---
title: "PublicGenoPheno AllTheBacteria Enterobacter"
output: html_document
date: "2024-10-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'png', echo = TRUE, message = FALSE, warning = FALSE,
                      cache = FALSE, fig.align = 'center')
knitr::opts_knit$set(root.dir = getwd())

# Data wrangling
library(tidyverse)
library(dplyr)
library(patchwork)
library(AMR)

source("../functions.R")
```


# read ATB AMRfinderplus calls - filter to high quality
```{r }

afp <-read_tsv("../AllTheBacteria/ATB_Enterobacter_AFP.tsv.gz") %>% filter(HQ)

denominator <- afp %>% select(Name) %>% unique() %>% nrow()

dim(afp)

denominator


enterobacter_complexes <- microorganisms.groups %>% filter(grepl("Enterobacter ", mo_name))

# one complex described - "Enterobacter cloacae complex"
table(enterobacter_complexes$mo_group_name)

# member species
enterobacter_complexes$mo_name

enterobacter_complexes <- enterobacter_complexes %>% 
  rename(complex=mo_group_name, Species=mo_name) %>% 
  select(complex, Species)

# add complex 
afp <- left_join(afp, enterobacter_complexes)
```

# check for core genes
``` {r}

# gene frequency per species
species_count <- afp %>% group_by(Name, Species) %>% count() %>% distinct() %>% ungroup() %>% group_by(Species) %>% summarise(nspp=n()) %>% arrange(-nspp)

gene_counts <- afp %>% group_by(Name, `Gene symbol`, Class, Subclass, Species, complex, `Element type`) %>% count() %>% distinct() %>% ungroup() %>% group_by(`Gene symbol`, Class, Subclass, Species, complex, `Element type`) %>% count() %>% left_join(species_count, by="Species") %>% mutate(freq=n/nspp)

write_tsv(gene_counts, file="Enterobacter_gene_freq_per_species.tsv")

node_counts <- afp %>% group_by(Name, `Gene symbol`, Class, Subclass, Species, complex, `Element type`, `Hierarchy node`) %>% count() %>% distinct() %>% ungroup() %>% group_by(`Hierarchy node`, Class, Subclass, Species, complex, `Element type`) %>% count() %>% left_join(species_count, by="Species") %>% mutate(freq=n/nspp)

write_tsv(node_counts, file="Enterobacter_node_freq_per_species.tsv")

```

# how is blaACT being called
``` {r}
afp %>% filter(grepl("blaACT",`Gene symbol`)) %>% select(Name) %>% unique() %>% nrow()

blaact_alleles <- afp %>% filter(grepl("blaACT",`Gene symbol`)) %>% group_by(`Gene symbol`, `Hierarchy node`, Subclass) %>% count() %>% arrange(-n)

nrow(blaact_alleles)
blaact_alleles %>% group_by(Subclass) %>% count()


# identify parent nodes for ACT genes
hierarchy <- read_tsv("../ReferenceGeneHierarchy.txt")
hierarchy %>% filter(node_id %in% blaact_alleles$`Gene symbol`) %>% group_by(parent_node_id) %>% count()

blaact_alleles %>% left_join(hierarchy, join_by("Hierarchy node"=="node_id"))  %>% view() %>% group_by(parent_node_id) %>% summarise(n=sum(n))

# tabulate blaACT hit combinations - they are independent
afp %>% filter(grepl("blaACT",`Hierarchy node`)) %>% left_join(hierarchy, join_by("Hierarchy node"=="node_id")) %>% select(Name, parent_node_id) %>% pivot_wider(names_from="parent_node_id", values_from = "parent_node_id", values_fn=max) %>% group_by(`CMY2-MIR-ACT-EC`, `blaACT`) %>% count()

parent_node_counts <- afp %>% left_join(hierarchy, join_by("Hierarchy node"=="node_id")) %>% group_by(Name, `Gene symbol`, Class, Subclass, Species, complex, `Element type`, `Hierarchy node`, parent_node_id) %>% count() %>% distinct() %>% ungroup() %>% group_by(parent_node_id, Class, Subclass, Species, complex, `Element type`) %>% count() %>% left_join(species_count, by="Species") %>% mutate(freq=n/nspp)

write_tsv(parent_node_counts, file="Enterobacter_parent_node_freq_per_species.tsv")

# plot parent nodes for ACT genes with per-species frequency

parent_node_counts %>% 
    filter(parent_node_id %in% c("CMY2-MIR-ACT-EC", "blaACT")) %>%
    mutate(label=paste0(Species, " (n=", nspp, ")")) %>%
    arrange(-nspp) %>%
    ggplot(aes(y=label, x=freq, fill=parent_node_id)) + geom_col(position="stack")

ggsave(width=6, height=6, file="Enterobacter_blaACT_bySpecies.pdf")

```

``` {r}

# check accessions associated with 'Gene symbol' of core genes
afp_core <- afp %>% filter(`Hierarchy node` %in% (node_counts %>% filter(freq>0.9 & `Element type`=="AMR") %>% pull(`Hierarchy node`))) 

# boxplot of identity of blaACT hits, stratified by accession
afp_core %>% 
  mutate(label=paste(`Gene symbol`, `Accession of closest sequence`)) %>% 
  filter(`Hierarchy node`=="blaACT") %>%
  ggplot(aes(x=label, y=`% Identity to reference sequence`)) + geom_boxplot() + coord_flip()

ggsave(width=5, height=9, file="blaACT_identity.pdf")


afp_core %>% 
  mutate(label=paste(`Gene symbol`, `Accession of closest sequence`)) %>% 
  filter(`Hierarchy node`=="blaACT") %>%
  ggplot(aes(x=label, y=`% Identity to reference sequence`)) + 
  geom_boxplot() + 
  coord_flip() +
  facet_wrap(~complex)

ggsave(width=8, height=9, file="blaACT_identity_byComplex.pdf")


afp_core %>% 
  mutate(label=paste(`Gene symbol`, `Accession of closest sequence`)) %>% 
  filter(`Hierarchy node`=="blaACT") %>%
  ggplot(aes(x=label, y=`% Identity to reference sequence`)) + 
  geom_boxplot() + coord_flip() +
  facet_wrap(~Species) + theme(axis.text =element_blank())

ggsave(width=12, height=9, file="blaACT_identity_bySpp.pdf")


```
```

# other core genes
``` {r}
# AMR markers with per-species frequency >90% (candidate core genes)
gene_counts %>% filter(nspp>10 & freq>0.9 & `Element type`=="AMR") %>% view()

# plot AMR markers with per-species frequency >90% (candidate core genes)
gene_counts %>% filter(nspp>10 & freq>0.9 & `Element type`=="AMR") %>%
  filter(!(`Gene symbol` %in% c("blaACT", "ampC"))) %>%
  mutate(label=paste0(Species, " (n=", nspp, ")")) %>%
  arrange(-nspp) %>%
  ggplot(aes(y=label, x=freq, fill=complex)) + 
  geom_col(position='dodge') + 
  facet_wrap(~`Gene symbol`) + theme(legend.position="bottom")

ggsave(width=10, height=10, file="Enterobacter_candidate_coreGenes.pdf")
```


# read AST data - from NCBI
```{r }

ast<-read_delim("../AST/AST_Enterobacter.tsv.gz") %>% rename(BioSample=`#BioSample`)

length(unique(ast$BioSample))

ast %>% group_by(BioSample) %>% summarise(n=length(unique((Antibiotic))))%>% pull(n) %>% summary()

```


# get AMRfinderplus and AST data where both are available
``` {r}
sum(unique(ast$BioSample) %in% afp$Name[afp$HQ])

# AST data for strains with AMRfinder results
ast_matched <- ast %>% filter(BioSample %in% afp$Name[afp$HQ])

# AMRfinder results, for strains with matching AST data from NCBI
afp_matched <- afp %>% filter(Name %in% ast$BioSample & HQ)

# check objects
dim(ast_matched)
ast_matched %>% pull(BioSample) %>% unique() %>% length()

dim(afp_matched)
afp_matched %>% pull(Name) %>% unique() %>% length()

# species
afp_matched %>% select(Name, Species) %>% distinct() %>% group_by(Species) %>% count()
```

# Enterobacter hormaechei_A
``` {r}
afp_eh <- afp_matched %>% filter(Species=="Enterobacter hormaechei_A")
ast_eh <- ast_matched %>% filter(BioSample %in% afp_eh$Name)
```

# solo markers for ciprofloxacin
``` {r}
cip_quinolone <- solo_marker_atb(ast_matched = ast_eh, afp_matched=afp_eh %>% filter(`Gene symbol`!="oqxB"), antibiotic = "ciprofloxacin", refgene_class = "QUINOLONE")

cip_quinolone$solo_stats

# excluding oqxAB as this is core
cip_quinolone_noOqx <- solo_marker_atb(ast_matched = ast_eh, afp_matched=afp_eh %>% filter(!(`Gene symbol` %in% c("oqxB", "oqxA"))), antibiotic = "ciprofloxacin", refgene_class ="QUINOLONE")

cip_quinolone_noOqx$solo_stats
cip_quinolone_noOqx$combined_plot

```

``` {r}
cip_binary <- getBinMatrix(ast_matched = ast_eh, afp_matched = afp_eh %>% filter(!(`Gene symbol` %in% c("oqxB", "oqxA"))), antibiotic = "ciprofloxacin", refgene_class = "QUINOLONE", maf=5)

cip_model <- logistf(resistant ~ ., data=cip_binary)

cip_logreg <- logreg_details(cip_model) %>% 
  mutate(sig=if_else(pval<0.05, TRUE, FALSE)) %>%
  filter(marker != "(Intercept)") %>%
  mutate(marker=gsub("`","",marker)) %>%
  ggplot(aes(y=marker, col=sig)) + 
  geom_vline(xintercept=0) + 
  geom_linerange(aes(xmin=ci.lower, xmax=ci.upper)) + 
  geom_point(aes(x=est)) + 
  scale_color_manual(values=c("grey", "black")) + 
  theme_light() + labs(x="Logist regression coefficient (95% CI)", col="p<0.05")

cip_logreg

```

# no AST data on fos
``` {r}
ast_eh %>% filter(Antibiotic=="fosfomycin") %>% nrow()
```

